Data Specialization for a Staged Lambda Calculus

### 2015-06-24 paper/semantics.tex
working on semantics section

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which is a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations that may contain second stage subterms with $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations that may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm,
and will be discussed in \ref{sec:needGround}.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement $\steptwo e {e'}$.
Since no reduction of stage two terms is to take place, it does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable,
as they would be in the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
For example, consider:
\begin{lstlisting} 
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $\hat{y}$, and substituting that variable instead:
\begin{lstlisting} 
2`let val yhat = e in prev{`1`[next{`2`yhat`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a hoisting step.
They steps are introduced when a \next\ block has contents that are a residual but not already a variable.
In essence, they operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.

>>>>

One of the central concepts in the formalization of \lang\ is the notion of a \emph{world},
which is a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second stage subterms with $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which that may not.
The distinction between these two first-stage worlds is necessitated by our splitting algorithm
and will be discussed in \ref{sec:needGround}.

@@

\paragraph{World \bbtwo.} Steps at world \bbtwo\ are given by the judgement $\steptwo e {e'}$.
Since no reduction of stage two terms is to take place, this judgement does nothing but search for \prev\ blocks within $e$.
A world \bbtwo\ term is done evaluating when it has the form $\exv q$, where $q$ is a residual. 
To be a residual, a term must have no first stage subterms (equivalently, no $\prev$s), even within the body of a function or branches of a case.
This implies that $\stepsym \bbtwo$ must proceed underneath second stage binders.

@@

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgement, $\stepmix e {e'}$, searches for redexes and reduces them in place.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other kinds of steps there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated when partial values are substituted for variable,
as they would be in the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
For example, consider:
\begin{lstlisting} 
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev([\next~e/x]e')$, but this may potentially duplicate an
expensive computation $e$. Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $\hat{y}$, and substituting that variable instead:
\begin{lstlisting} 
2`let val yhat = e in prev{`1`[next{`2`yhat`1`}/x]e'`2`}`
\end{lstlisting}
This behavior is implemented by the $\lift e {e'}$ judgement, called a hoisting step.
These steps are introduced when a \next\ block has contents that are a residual but not already a variable.
In essence, they operate by ``bubbling up'' a subtitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
